---
title: "R_notes"
author: "Xian"
date: "2/23/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

library(broom)
library(car)
library(DescTools)
library(dplyr)
library(dummies)
library(GGally)
library(ggplot2)
library(gridExtra)
library(knitr)
library(lmtest)
library(lsr)
library(mosaic)
library(modelr)
library(orcutt)
library(olsrr)
library(plotly)
library(ppcor)
library(psych)
library(readxl)
library(reticulate)
library(survival)
library(tidyverse)
```


Contents:
1. Useful commands 
2. data types and convert
3. Semantics/Syntax of R
4. Latex type the symbols
5. Import/view/index data
6. Create/manipulate data frame
7. Save/Write data 
8. Visualize with plots 
9. Power analysis
10. Check model assumptions
11. Stats info (r/p/q/dnorm)
12. Stats tests (t/anova/ancova/corr/manova/ancova)
13. Effects (moderation/mediation)
14. Fit regression models (lm/glm/rob/stepwise/mixed)
15. Compare model fit (nested models)
16. Additional plots (ROC/Network)




===========================================================================
# Useful commands 

*create new chunk of R code:*
1) Command + Option + I
1) Ctrl + Option + I
2) mannually type in '''{r} and '''
*run current chunk of R code:*
1) Command + Shift + Enter
2) Command + Option + C
*comment off and back on:*
1) Command + Shift + C
1) Ctrl + Shift + C
*clear work space:*
1) rm(list = ls())
2) rm(nameOftheDForVar)
*help functions:*
1) ?round
2) help(round)
*math/calculate:*
- `abs()`: returns the absolute value of numeric objects
- `max()`: returns the largest element of a numeric vector
- `min()`: returns the smallest element of a numeric vector
- `sqrt()`: returns the square root of numeric objects
- `sum()`: computes the sum of the elements of a numeric vector
- `round()`: rounds a numeric object to a designated number of digits
*useful others*
- `install.packages('PackageName')`: install package
- `library(PackageName)`: load in the package
- `search()`: see loaded packages in the current env
- `getwd()`: see current working directory
- `setwd('ThePathYouWant')`: set the wd to another path
- `list.files()`: see what files are available from the current wd
- `who()`: names and describes the objects in the workspace



===========================================================================
# R data types and converting data type

```{r}
# numeric
a1 = 12 # as.numeric() changes data type to num
a2 = c(1,1,2,3,5,8,13) #combines elements into a vector
a2
# character
b1 = toupper('the top') #change char to upper cases; tolower for lower cases
b2 = '12'# as.character() changes data type to char
b1
# factor
c1 = factor(b1) 
c2 = factor(c(a1,a2,b1,b2)) #auto show all existing levels of the factor, but can also check using: levels() 
c2
# logical
d1 = TRUE 
d2 = (a1 == a2) #compares a1 with each of the elements in a2
d3 = (length(a2) == nchar(b1)) #length() returns #elements; nchar() returns #of char within this element
d3
# data type:
class(a1);class(a2) #get the class of the data, i.e. data type in R
class(b1);class(b2)
class(c1);class(c2)
class(d1);class(d2)
```



===========================================================================
# Semantics/Syntax of R:

while/for/ifelse/function
```{r}
# switch 
#loop through the list until find the match for the expression, return the matched value.
#If the value evaluated is a number, that item of the list is returned.
switch(2,"red","green","blue") 
switch(4,"red","green","blue")
#if expression is str, the matching named item’s value is returned
switch("color", "color" = "red", "shape" = "square", "length" = 5) 
switch("length", "color" = "red", "shape" = "square", "length" = 5)


# while loop
dice1 = c(1:6)
dice2 = c(1:6)
roll = c(sample(dice1,1),sample(dice2,1))
#roll the two dices until they sum to 7
while(sum(roll)!=7) {
  roll = c(sample(dice1,1),sample(dice2,1))
  print(sum(roll))}
#flip coin until all three are Heads
coin = c("Heads", "Tails")
flip = rep("Tails", 3) #repeat 3 times
while(any(flip != "Heads")) {
  flip = sample(coin, size = 3, replace = TRUE)
  print(flip)}


# for loop
for(i in 1:5) {
  print(i + i)}
#assign letter grades based on numeric points scored
score <- c(73,68,81,95,74,85,92,60,77,88)
letter <- rep(NA, 10)
grades <- data.frame(score, letter)
for(i in 1:nrow(grades)) {
  if(grades$score[i] >= 90){
    grades$letter[i] <- "A"
  } else if(grades$score[i] < 90 & grades$score[i] >= 80) {
    grades$letter[i] <- "B"
  } else if(grades$score[i] < 80 & grades$score[i] >= 70) {
    grades$letter[i] <- "C"
  } else if(grades$score[i] < 70 & grades$score[i] >= 65) {
    grades$letter[i] <- "D"
  } else {grades$letter[i] <- "F"}} 
grades
# create a matrix of data and compute a new variable as the mean of every row
set.seed(100)
datamat <- matrix(data = round(runif(n = 25, min = 1, max = 7)), nrow = 5, ncol = 5)
datamat <- cbind(datamat, avg=rep(NA, 5))
for(i in 1:nrow(datamat)) {
  datamat[i, "avg"] <- mean(datamat[i, 1:5])}
datamat


# if/(else if)/else
if(2 > 2) {
  print("something doesn't seem right")
} else if(2 < 2) {
  print("this doesn't seem any better")
} else {
  print("it would've been easier to go with '2 == 2'")}
#The `ifelse` function consolidates the condition itself with the comands that should execute if it is TRUE vs. FALSE
ifelse(test = 1 + 1 == 2, yes = "correct", no = "wrong")
# if/else &func
NHST = function(smp.mean, pop.mean, pop.std, smp.size) {
  z.score = (smp.mean-pop.mean)/(pop.std/sqrt(smp.size))
  p.value = 1-pnorm(abs(z.score),0,1)
  tail = 2
  alpha = .05/tail
  decision = 0
  if(p.value>alpha) {decision= 'retain null hypothesis'} 
  else{decision= 'reject null hypothesis'}
  print(z.score)
  print(p.value)
  print(decision)} 
#input sample mean&size, pop mean&std; returns z/p/decision
set.seed(12)
data = rnorm(100, 16.5, 4.5)
NHST(mean(data),15,8.5,length(data))
NHST(mean(data),19,2.5,length(data))


# Function
plotreg = function(data, x, y, xname, yname) {
  library(ggplot2)
  graph <- ggplot(data = data, mapping = aes(x = x, y = y)) +
    geom_point(color = "dark blue", position = "jitter") +
    geom_smooth(method = "lm", color = "dark red", fill = "dark red", alpha = .1) +
    labs(x = xname, y = yname) +
    theme_bw()
  return(graph)} 
#Currently, our custom functions only exist in our active session of *R*
#To retain them, we must save them as ".R" files:
dump("plotreg", file = "plotreg.R")
#Now, we can remove the `plotreg` function from our workspace 
rm(plotreg)
#and summon it through its saved source code
source("plotreg.R")
```



===========================================================================
# Latex type the symbols:

$-\infty$
$\sigma$
$\beta$
$\beta_1$
$\beta^2$
$\beta^{1/2}$
$\hat{Y}$
$\hat{\sigma}^2$
$H_a:\space$
$1\over{Cov(X,Y)}$
$R^2=1-{SSE\over{SST}}$
$\hat{\sigma}=\sqrt{\frac{SSE}{n-p-1}}$
$$H_0: \beta_0=\beta_1 \\ H_a:\beta_0 \neq \beta_1$$
$$\hat{b}_0=\bar{X}_G\qquad \hat{b}_1=\frac{(\frac{\bar{X}_2+\bar{X}_3}{2})-\bar{X}_1}{3}\qquad \hat{b}_2=\frac{\bar{X}_3-\bar{X}_2}{2}$$


===========================================================================
# Import/view/index data:

```{r}
library(car) 
library(readxl) 

# read in txt/csv/excel
States = read.delim("example_input/data_P152.txt") #can also read in http://www.../data.txt
States2 = read.csv("example_input/data_P152.csv",header = TRUE) 
States3 = read.table("example_input/data_P152.csv",header = TRUE,sep = ",") 
cd1 = read.table("example_input/data_classdata1.txt", header = TRUE)
RT = read_excel("example_input/data_rt.xlsx",1) #the first sheet

# view data
States
States2
States3
head(States) #shows first several rows of the df
str(States) #structure of the data, variable-wise
dim(States) #check dataframe dimension (nrows ncols)
summary(States) #data summary of min, max, etc.
describe(cd1) #more detailed descriptive summary

# view data descriptives
Freq(cd1$TimeGrp) #check the frequency of each speed category
mean(cd1$Time)
mean(cd1$Time, trim = .10) #trim the top 10% and buttom 10% of the data before calculating mean
median(cd1$Time)
mode(cd1$Time)
library(lsr)
modeOf(cd1$Time)
range(cd1$Time)
diff(range(cd1$Time))
quantile(cd1$Time)
IQR(cd1$Time)
var(cd1$Time)
sd(cd1$Time)
sqrt(var(cd1$Time))
describe(cd1$Time, IQR = TRUE)
describe(winsor(cd1$Time, trim = .10))

# index data
States[,2] #index all rows, 2nd col; returns a vector 
States[3,2] #index the 3rd row, 2nd col; returns a vector
States[4,] #index the 4th row, all cols; returns a df
States[1:3,c(2,3,7)] #index row 1-3, col 2,3,7; returns a df
which(States$STATE=='AK') # check the index of 'AK' from the variable 'STATE'
States$STATE[49] #print the data of this index to check if it's 'AK'
States$STATE[States$STATE=='AK'] # alt to above, directly index the one with 'AK'; returns 'AK' if it's in the var
```



===========================================================================
# Create/manipulate data frame:

Create Dataframe
```{r}
# 1) hand-create dataframe
chocolate <- 37
vanilla <- 58
mint.chip <- 25
ice.cream <- cbind(chocolate, vanilla, mint.chip)
rownames(ice.cream) <- "Observed" #assign row name 
colnames(ice.cream) <- c("Chocolate", "Vanilla", "Mint Chip")
ice.cream

# 2)
States = read.delim("example_input/data_P152.txt")
# getting vectors of num
y = States$Y
x = States$X
i = c(1:length(x)) 
y_dev = y - mean(y)
x_dev = x - mean(x)
y_var = y_dev^2
x_var = x_dev^2
cov_xy = x_dev*y_dev
# getting data-frame (tibble better than df)
mydata = tibble(i,y,x,y_dev,x_dev,y_var,x_var,cov_xy)
mydata

# 3) 
# getting vectors of num and factor
height = c(61, 49, 55, 53, 63, 58)
points = c(24, 88, 12, 49, 64, 39)
cond= factor(c(1, 1, 2, 2, 1, 2), labels = c("dominant", "non-dominant"))
# getting data-frame from vector
bball = data.frame(condition, height, points)
bball

# 4) 
# getting mat from vectors
v2= cbind(i,y,x,y_dev,x_dev,y_var,x_var,cov_xy)
# convert mat to tibble
newData.tb = as_tibble(v2)
newData.tb

# 5)
# getting matrix from nums
v1=matrix(c(0.32,6,0,1,0,
            0.42,5,0,0,0,
            0.50,4,0,1,0,
            0.62,3,1,0,0,
            0.80,6,0,1,0,
            0.87,5,1,0,1,
            0.98,4,0,0,0),ncol=5,byrow=TRUE)
# convert mat to df
newData.df = as.data.frame(v1)
# name each col for the new df
names(newData.df) = c("carat","numcolor","VS","VVS","fair")
newData.df
```

Manipulate Data
```{r}
# using dplyr
library(dplyr)
library(nycflights13)

# dplyr::select
# select the cols/vars you want; reorder the cols
flights %>%
  select(arr_time, month:dep_delay)
flights %>% # reorder the cols with arr_time up front
  select(arr_time, everything()) # everything() calls all unmentioned vars
flights %>%
  select(contains("time")) %>% #use contains()
  rename(depart = dep_time,
         arrive = arr_time)

# dplyr::filter/top_n
# apply certain criteria and extract only rows of data that fit those criteria
flights %>%
  filter(dest == "PDX") # get rows only flying to portland:
flights %>%
  filter(origin == "JFK", (dest == "BTV" | dest == "SEA"), month >= 10) # multiple restrictions
flights %>%
  select(arr_delay) %>%
  top_n(n = 5, wt = arr_delay) #get the top 5 arr_delay rows 

# dplyr::summarize
# computing summary statistics from the raw data using functions, resulting in a small df
weather %>%
  summarize(mean = mean(temp, na.rm = TRUE), std_dev = sd(temp, na.rm = TRUE)) 
weather %>%
  dplyr::summarize(count = n(), sweltering = max(humid, na.rm = TRUE), soaked = max(precip, na.rm = TRUE))

# dplyr::group_by
# segments any subsequent output into levels of a factor
weather %>%
  group_by(month) %>% # group by a var when summarizing stats
  dplyr::summarize(count = n(), # count data-entry/rows for the groupped var
            mean = mean(temp, na.rm = TRUE),
            std_dev = sd(temp, na.rm = TRUE))
flights %>%
  group_by(origin, carrier) %>% # group by two vars
  dplyr::summarize(count = n())

# dplyr::arrange
# sort data in ascending or descending order based on values on one or more var
flights %>%
  arrange(desc(gain)) #desc: orange in descending order; default is ascending order
flights %>%
  group_by(dest) %>%
  dplyr::summarize(num_flights = n()) %>%
  arrange(desc(num_flights))

# dplyr::mutate
# create new variables within a data set by manipulating existing values
flights %>%
  mutate(hours = air_time / 60,
         gain_per_hour = gain / hours)

# dplyr::join
# `inner_join`, one of several "join" functions within `dplyr`: merge two data frames into one
flights %>%
  inner_join(airlines, by = "carrier") # join by the var shared between both datasets
flights %>%
  inner_join(airports, by = c("dest" = "faa")) #dest is in airdata and faa is the same thing in df airports
flights %>%
  inner_join(weather, by = c("year", "month", "day", "hour", "origin")) #for multiple shared vars
#bind dataframe by row: rbind(df1,df2)

# dplyr + ggplot
flights %>%
  group_by(month) %>%
  dplyr::summarize(delay = mean(arr_delay, na.rm = TRUE)) %>%
  ggplot(aes(x=month, y= delay))+
  geom_point()+
  geom_line(group=1)

# dplyr::gather
# change the df from wide format to long format, keeping the pID column
library(tidyr); library(dplyr)
rmwide = read.table("example_input/rm.dat", header = TRUE)
rmlong = rmwide %>% gather(key = "cond", value = "search", -pID) #change wide to long format
rmwide; rmlong
```


```{r}
# 1) filter data
States = read.delim("example_input/data_P152.txt")
States
States[States$X>5,c(0:-1)]
States[1:nrow(States),c('STATE','X','X1','Region')] #select all rows and specified columns
subset(States, Y>=180 & X>10, select=c(STATE,X,X1,Region)) # select a subset of dataframe
States=States %>% filter(STATE!="AK") %>% droplevels() # keeps the non-'AK'
States$STATE[49] # check if this index has changed; it has (coz 'AK' has been dropped) 
States$STATE[States$STATE=='AK'] # another way to check if 'AK' is still in the data
TD= RT %>% filter(group=="0") # keeps group0 
DP= RT %>% filter(group=="1") # keeps group1
head(DP)
# drop missing values from dataframe
States$Y[States$Y<=180] = NA # set all the Y smaller than 180 to be NA
States$Y #checking that there are indeed NA in the var Y of df States 
subset(States, !is.na(Y), select=c(STATE,Y,X,X1,Region)) #subset, row selection restricts to non-NA
States %>% filter(!is.na(Y)) #keep the non-NA in, leave out the NA
States[complete.cases(States),] #selected rows that has TRUE to 'complete case', i.e. non-NA
# drop missing values from vector
vector = c(2, 4, NA, 7) #create a vector with missing value
mean(vector) #mean cannot be calculated due to NA
mean(vector, na.rm = TRUE) # calculate the mean by rm na
vector[!is.na(vector)] #drop missing value
vector[complete.cases(vector)] #a 2nd way to drop missing value
na.omit(vector) #a 3rd way to drop missing value 
na.exclude(vector) #a 4th way to drop missing value 


# 2) sort data
attach(mtcars)
# sort by mpg
newdata <- mtcars[order(mpg),] 
# sort by mpg and cyl
newdata <- mtcars[order(mpg, cyl),]
#sort by mpg (ascending) and cyl (descending)
newdata <- mtcars[order(mpg, -cyl),] 
detach(mtcars)
newdata


# 3) cut num into categories
cd1 = read.table("example_input/data_classdata1.txt", header = TRUE)
cd1$Time
#create another var 'TimeGrp' and divide time into several speed categories (setting the range for the cut)
cd1$TimeGrp = cut(cd1$Time, breaks = c(35, 50, 65, 80, 95, 110, 125), labels = c("Fastest", "Fast", "Kinda Fast", "Slow-ish", "Slow", "Very Slow"))
cd1$TimeGrp
# create 2 size categories 
newdata$newvar <- ifelse(newdata$drat > 3.5, 
c("larger"), c("smaller")) 
newdata
# create 3 size categories 
newdata$agecat[newdata$drat > 4] <- "largest"
newdata$agecat[newdata$drat > 3 & newdata$drat <= 4] <- "Middle sized"
newdata$agecat[newdata$drat <= 3] <- "smallest"
newdata
#create dummy variables
library(dummies)
letters = c( "a", "a", "b", "c", "d", "e", "f", "g", "h", "b", "b" ) 
dummy( letters)
cd1 = read.table("example_input/data_classdata1.txt", header = TRUE)
mat = dummy(cd1$Stim)
stim1 = mat[1:nrow(mat),1]
stim2 = mat[1:nrow(mat),2]
stim3 = mat[1:nrow(mat),3]
cd1 %>% mutate(stim1_dummy=stim1,stim2_dummy=stim2,stim3_dummy=stim3)


# 4) convert nominal to ordinal 
DD=read.delim("example_input/data_diamonds.txt")
head(DD)
# create another two cols of var to add in df: one for VS (0=not VS, 1= is VS), one for VVS  
DD=DD%>%mutate(VS=1*(clarity=="VS1")+1*(clarity=="VS2"),VVS=1*(clarity=="VVS1")+1*(clarity=="VVS2")) 
summary(DD)
# set the conversion criteria: what shall replace what
conversion=c("D"=6,"E"=5,"F"=4,"G"=3,"H"=2,"I"=1,"J"=0)
# use this conversion on a var; assign it as another var to df 
DD$numcolor=conversion[DD$color] 
DD=DD%>%mutate(fair=1*(cut=="Fair"))
DD

# fit a linaer reg model using dimond data from DD  
fit=lm(price~carat+numcolor+VS+VVS+fair,data=DD)
library(broom)
tidy(fit)
# get fitted values and errors
fitted= as.integer(fitted(fit))
errors = as.integer(DD$price-fitted(fit))
DD = DD%>% mutate(fitted, errors)
head(DD)
# predict the prices of dimond data from newData.df, using the above fitted lm
predicts = predict(fit,newData.df) #need to run previous block to read in newData.df 
predicts

```



===========================================================================
# Save data

save dataframe to csv/txt/xls files
```{r}
# write df as csv
write.csv(mydata,"example_output/states_df.csv", row.names = TRUE)
# write df as txt
write.table(DD, "example_output/dimond_fit_df.txt", append = FALSE, sep = ",", dec = ".", row.names = TRUE, col.names = TRUE)
# write df as xlsx
library("writexl")
write_xlsx(newData.df,"example_output/dimond_test_df.xlsx")
```

save figure to .png/.tif/.jpg
```{r}
States = read.delim("example_input/data_P152.txt")
gf_point(Y~X1,data=States) %>% gf_smooth(method="lm")
ggsave('example_output/name_of_the_plot.jpg', width = 5, height = 3)
```


===========================================================================
# Visualize with plots:

*gf_points and ggpair:*
```{r}
States = read.delim("example_input/data_P152.txt")

#gf_point
p1= gf_point(Y~X1,data=States) %>% 
  gf_smooth(method="lm")

p2= gf_point(Y~X2,data=States) %>% 
  gf_smooth(method="lm",se=TRUE,linetype=1,color="yellow") %>%
  gf_labs(x="predictor X2",y="response Y")

p3= gf_point(Y~X3,data=States) %>% 
  gf_smooth(method="lm",se=TRUE,linetype=2,color="red") %>%
  gf_labs(x="predictor X3",y="response Y")

p4= gf_point(X1~X2,data=States, color = 'brown') %>% 
  gf_smooth(method="lm",se=TRUE,linetype=3,color="black")%>%
  gf_labs(x="predictor X2",y="predictor X1")

p5= gf_point(X2~X3,data=States) %>% 
  gf_smooth(method="lm",se=TRUE,linetype=4,color="purple")%>%
  gf_labs(x="predictor X3",y="predictor X2")

p6= gf_point(X1~X3,size=~Region,color=~Region,data=States) %>% 
  gf_smooth(method="lm",se=TRUE,linetype=5)%>%
  gf_labs(x="predictor X3",y="predictor X1")

library(gridExtra)
grid.arrange(p1,p2,p3,p4,p5,ncol=3)
p6

# arrange multiple graphs in one (example) and save
conn = read_excel("example_input/data_conn.xls", sheet = 4)
p1= gf_point(chg_score ~ reading,data=conn,xlab="SPL&SMC FC increase", ylab="score increase (post-pre)",
             subtitle="partial.r.sq=0.14") %>% gf_smooth(method="lm")
p2=gf_point(chg_score ~ listening,data=conn,xlab="listening", ylab="score increase (post-pre)",
            subtitle="partial.r.sq=0.33") %>% gf_smooth(method="lm")
p3=gf_point(chg_score ~ speaking,data=conn,xlab="speaking", ylab="score increase (post-pre)",
            subtitle="partial.r.sq=0.28") %>% gf_smooth(method="lm")

grid.arrange(p1,p2,p3,ncol=3,
              top="Relationship Between the Change of FC & Score (Post-Pre)",
              bottom="model r.squared=0.40, adj.r.sq=0.34")

# ggpairs
library(GGally)
ggpairs(States%>%select(3:6), lower=list(continuous=wrap("smooth",method="lm",color="blue")),progress=F)
ggpairs(States[,c(1,3,4,5,6)],lower=list(continuous=wrap("smooth",method="lm",color="blue")),progress=F)
```

*ggplots & defaultPlots:*
```{r}
RT = read_excel("example_input/data_rt.xlsx",1)

library(ggforce) #for geom_sina
# ggplots for point/line/bar/hist/boxplot/jitter/violin/sina
p1 = ggplot(RT, aes(factor(group), ffmt)) + geom_point()
p2 = ggplot(RT, aes(factor(group), ffmt)) + geom_boxplot()
p3 = ggplot(RT, aes(group, ffmt))         + geom_line()  #geom_vline, geom_hline for vertical/horizontal lines
p4 = ggplot(RT, aes(factor(group), ffmt)) + geom_jitter()
p5 = ggplot(RT, aes(factor(group), ffmt)) + geom_violin()
p6 = ggplot(RT, aes(factor(group), ffmt)) + geom_sina()
p7 = ggplot(RT, aes(factor(group)))       + geom_bar()
p8 = ggplot(data=RT, aes(x=cfmt))         + geom_bar()
p9 = ggplot(data=RT, aes(x=cfmt))         + geom_histogram(color='black',fill='red',alpha=1/2)

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,ncol=3)

p1 = ggplot(RT, aes(factor(group), ffmt)) + geom_violin() + geom_sina()
p2 = ggplot(RT, aes(factor(group), ffmt)) + geom_sina(aes(color=factor(group)))
p3 = ggplot(RT, aes(factor(group), ffmt)) + geom_violin(alpha = 1/10) + geom_sina(aes(color=factor(group))) 
p4 = ggplot(RT, aes(x= benton_acc, y = com, color = group) ) + geom_point() +
     geom_smooth(method = "lm", alpha = .15, se = TRUE)
grid.arrange(p1,p2,p3,p4,ncol=2)

# alternatives
hist(RT$ffmt,col="chocolate",breaks=15) #for p9 
plot(RT$cfmt,RT$ffmt)
scatterplot(ffmt~cfmt,data=RT)
library(sinaplot) #for p6
n_group = length(unique(RT$group))
sinaplot(ffmt~group, data=RT, pch = 20, col = rainbow(n_group), method = "density", adjust = 1/20)
gf_point(ffmt~benton_acc,color=~group, data=RT) %>% # for p4
  gf_smooth(method="lm",se=TRUE,linetype=1,color="blue",group=~group) 

```

*ggplots & plotly:*
```{r fig.height=5, fig.width=10}
library(ggplot2) 
diamonds[1:6, c(1:5, 7)] #diamond is one of the default R datasets
set.seed(100) #get a random set of 100 numbers and fix this set (don't change)
dia = diamonds[sample(nrow(diamonds), size = 100), ] #get a subset of diamond df (100 rows)
dia

# 1). 
# Scaterplots
#plotting the scatter of carat and price
ggplot(data = dia, mapping = aes(x = carat, y = price)) + #map the var to aes properties of the plot
  geom_point(mapping = aes(color = cut), shape = "triangle") #only when want to map var value to points do we use aes
ggplot(data = dia, mapping = aes(carat, price)) +
  geom_point(aes(color = cut, size = clarity), shape = 18) #adding addition properties to the points
#adding another layer: the trendline divided by cut with different color (in full range)
ggplot(data = dia, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = cut, size=clarity), size = 5) +
  geom_smooth(method = "lm", se = FALSE, aes(color = cut), fullrange=TRUE,linetype='twodash') 
#adding another layer: the labels 
ggplot(data = dia, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), size= 5) +
  geom_smooth(method = "lm", color= 'black', se = FALSE, aes(lty = cut)) + #lty: line-type
  labs(title = "Diamond Quality Scatterplot", x = "Carat Size", y = "Price in Dollars")
#avoid overplotting 
#(too many points in one graph so many points may overlay each other): **jittering** and **facetting**
ggplot(data = diamonds[diamonds$price < 500, ], aes(x = carat, y = price)) +
  geom_point(color = "blue") #e.g. many points overlay each other in this plot
#this is **jitter**: jitter the points so that in graph, they slightly move position away from each other
ggplot(data = diamonds[diamonds$price < 500, ], aes(x = carat, y = price)) +
  geom_point(color = "blue", position = "jitter") 
#this is another jitter 
ggplot(data = diamonds[diamonds$price < 500, ], aes(x = carat, y = price)) +
  geom_jitter(color = "blue")
#this is **Facet wraps**: create separate plots at each level of a designated variable
ggplot(data = diamonds[diamonds$price < 500, ], aes(x = carat, y = price)) +
  geom_point() +
  facet_wrap(~ cut)
#this is **Facet grids**: create separate plots at every pairing of levels of 2 designated variables
ggplot(data = diamonds[diamonds$price < 500, ], aes(x = carat, y = price)) +
  geom_point(aes(color = depth), shape='triangle', size=1) +
  facet_grid(clarity ~ cut) 

# 2) 
# Line graph / Line plot
data = read_excel('example_input/sub2.xlsx'); data
data$SemanticCent = as.numeric(data$SemanticCent) 
plot(data$event,data$SemanticCent,type='b') #both x and y are continuous
ggplot(data,aes(x=event,y=SemanticCent))+
  geom_point(size=4,color='orange')+
  geom_line(color='orange',lwd=3)
# another example
allsub = read.table('example_input/allsub_results.csv',header=TRUE,sep=',')
allsub$sub.num=as.factor(allsub$sub.num)
event = sort(unique(allsub$event))
ggplot(allsub, aes(x=event,y=caus.z))+
  geom_point(aes(color=sub.num))+
  geom_line(aes(color=sub.num))+
  scale_x_continuous(breaks=seq(1, max(allsub$event),2))+
  scale_y_continuous(breaks=seq(-10, 10, 0.5))+
  theme(axis.text.x = element_text(angle = 90))+
  geom_vline(xintercept = event,lwd=0.1,linetype=6,color='blue')
# with Errorbar
corr = read.table('example_input/corr_results.csv',header=TRUE,sep=',')
ggplot(corr, aes(x=sub.num,y=sem.caus.r))+
  geom_point()+
  geom_line(color='purple')+
  scale_x_continuous(breaks=seq(1,28,1))+
  geom_hline(yintercept = 0, lwd=2,alpha=0.3)+
  geom_hline(yintercept = mean(corr$sem.caus.r), lwd=2,alpha=0.3,color='purple')+
  geom_vline(xintercept=subjs,lwd=0.3,linetype=6,color='grey')+
  ylab('semantic-causal r')+
  xlab('subject number')+
  geom_errorbar(aes(ymin=sem.caus.95ci.low, ymax=sem.caus.95ci.high), width = 0.2,lwd=0.5)

# 3).
# Bar graph (ggplot stats)
# geom background computation for plotting called by stat_summary/stat_function from Hmisc
library(Hmisc)
#the default bar graph always has counts as y-axis
ggplot(diamonds, aes(x = price)) +
  geom_bar(aes(fill=cut))
#calling mean from the stat_summary function, setting geom to be bar (resets the y-axis to mean)
ggplot(diamonds, aes(x = cut, y = price)) +
  stat_summary(fun.y = mean, geom = "bar")
#adding additional layer (errorbar needs both SE and mean, so needs to be called from fun.data; so is confidence interval)
ggplot(diamonds, aes(x = cut, y = price)) +
  stat_summary(fun.y = mean, geom = "bar", aes(fill = clarity), position = 'dodge') +
  stat_summary(fun.data= mean_se, geom= "errorbar", width= .25, aes(group= clarity), position= position_dodge(width= .9)) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", aes(group= clarity), position = position_dodge(width= .9)) 
#adding additional layer of line (group by 1: one line through all points)
ggplot(diamonds, aes(x = cut, y = price)) +
  #stat_summary(fun.y = mean, geom = "bar", aes(fill = clarity), position = 'dodge') +
  stat_summary(fun.y = mean, geom = "point", size = 2, col = "dark blue") +
  stat_summary(fun.y = mean, geom = "line", aes(group = 1), lty = "dashed") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .2)

# 4). 
# Boxplots 
#boxplot for price (split by cut in filling color)
ggplot(dia, aes(y = cut)) + 
  geom_boxplot(aes(fill=cut)) 
#boxplot for price by cut (with outliers specify and axis flip)
ggplot(dia, aes(x = cut, y = price)) +
  geom_boxplot(aes(fill = cut),  outlier.shape = "o", outlier.size = 5)  + #, outlier.color = "red", show.legend = FALSE
  coord_flip()
#plot the range
allsub = read.table('example_input/allsub_results.csv',header=TRUE,sep=',')
allsub$sub.num=as.factor(allsub$sub.num)
library(dplyr)
df = allsub %>% group_by(event)%>% 
  summarize(count = n(), caus.z.mean = mean(caus.z, na.rm = TRUE), caus.z.min=range(caus.z,na.rm=TRUE)[1], caus.z.max=range(caus.z,na.rm=TRUE)[2])
ggplot(df, aes(x=event))+
  geom_linerange(aes(ymin=caus.z.min,ymax=caus.z.max),linetype=2,color="blue")+
  geom_point(aes(y=caus.z.min),size=3,color="red")+
  geom_point(aes(y=caus.z.max),size=3,color="red")+
  theme_bw()+
  ylab('Causal Centrality per event (range across subjects)')

# 5)
# Histograms
#histogram: visualize the shape of distributions of data on a var
ggplot(diamonds, aes(x = price)) +
  geom_histogram(color = "blue", fill = "gold")+ 
  theme_bw() # theme_classic() 
#histogram added another layer (the normal curve with data mean/std)
ggplot(diamonds, aes(x = price)) +
  geom_histogram(aes(y = ..density..), color = "blue", fill = "gold") +
  stat_function(fun = dnorm, args = list(mean = mean(diamonds$price), sd = sd(diamonds$price)), col = "dark red", lwd = 1.5) +
  theme_classic()
#another histogram with vlines for mean, median and mode
library(dplyr)
mode = (midwest %>% group_by(area) %>% summarise(count = n()) %>% arrange(desc(count)))$area[1]
ggplot(midwest, aes(x=area)) +
  geom_histogram(aes(y = ..density..), color='black', fill='pink', bins = 50) +
  stat_function(fun=dnorm,args= list(mean= mean(midwest$area), sd= sd(midwest$area)), color='purple', lwd=1.5) + 
  geom_vline(xintercept = c(mean(midwest$area), median(midwest$area), mode), color= c('green','yellow','blue'))

# 6)
# Areas under the curve
#this is to generate two normally distributed datasets with different mean and sd 
data1 = tibble(value = rnorm(1000, mean= 10, sd= 3))
#this is the alpha value with p<.05 for either a two-tailed test or a one-tailed test
alpha1 = qnorm(0.975,mean=10, sd=3)
alpha1.left = qnorm(0.025,mean=10, sd=3)
#this is for two-tailed test with p<.05 (so .025 on each tail); beta is the red shaded area 
ggplot(data1, aes(value), xlim=c(0,20)) + 
  stat_function(fun=dnorm,args=list(mean=10,sd=3),col="dark blue",lwd=1.5,xlim=c(0,25)) + # curve1
  stat_function(fun=dnorm,args=list(mean=10,sd=3),col="dark blue",lwd=1.5,xlim=c(alpha1,25),geom = 'area',fill ='dark blue',alpha=0.5) + # area1: alpha
  stat_function(fun=dnorm,args=list(mean=10,sd=3),col="dark blue",lwd=1.5,xlim=c(0,alpha1.left),geom ='area', fill ='dark blue',alpha=0.5) + # area2: alpha
  stat_function(fun=dnorm,args=list(mean=18,sd=2),col="dark red",lwd=1.5,xlim=c(0,25)) + # curve2
  stat_function(fun=dnorm,args=list(mean=18,sd=2),col="dark red",lwd=1.5,xlim=c(alpha1.left,alpha1),geom ='area',fill ='dark red',alpha = 0.5)+ # area2: beta
  stat_function(fun=dnorm,args=list(mean=18,sd=2),col="dark red",lwd=1.5,xlim=c(alpha1,25),geom='area',fill ='dark blue',alpha = 0.5)+ # area3: power
  geom_vline(xintercept = c( alpha1.left,alpha1), color='black',lwd=1) + #alpha threshold
  labs(title='Two-Tailed p<.05; beta = red shaded area') +
  theme_classic() 

# 7) 
# linking dots with lines
#this is to link the dots to a specified horizontal line segment
waldo = read.table("example_input/waldo.dat", header = TRUE)
ggplot(waldo, aes(ID, search)) +
  geom_segment(aes(x = 0, xend = 21, y = mean(search), yend = mean(search))) +
  geom_point(aes(col = cond)) +
  geom_segment(aes(x = ID, xend = ID, y = mean(search), yend = search), lty = 2, lwd = .5) +
  theme_classic()
#this is to plot dots to group means
ggplot(waldo, aes(ID, search)) +
  geom_segment(aes(x = 0, xend = 21, y = mean(search), yend = mean(search)), alpha = .25) +
  geom_point(aes(col = cond)) +
  geom_segment(aes(x = 1, xend = 6, y = mean(search[cond == "ctrl"]), yend = mean(search[cond == "ctrl"])), col = "red") +
  geom_segment(aes(x = 7, xend = 12, y = mean(search[cond == "info"]), yend = mean(search[cond == "info"])), col = "green3") +
  geom_segment(aes(x = 13, xend = 18, y = mean(search[cond == "time"]), yend = mean(search[cond == "time"])), col = "blue") +
  geom_segment(aes(x = c(1:18), xend = c(1:18), y = search[1:18], yend = c(rep(mean(search[cond == "ctrl"]),6), rep(mean(search[cond=='info']),6), rep(mean(search[cond=='time']),6))), lty = 2) +
  theme_classic()

# 8)
# plot matrix
library(plot.matrix)
mat = read.table('example_input/caus_super_matrix.csv',sep=',')
mat = as.matrix(mat)
mat[mat == 0] <- NA
caus.super.mat = mat
plot(caus.super.mat,border=NA,col=topo.colors,na.col="black",fmt.key="%.1f",xlab="event label", ylab="event label")

# 9)
# plotly
library(ggplot2)
library(plotly)
myplot= ggplot(dia, aes(x= carat, y= price)) + 
  geom_point(size=2, color = 'magenta')+
  geom_smooth(method="lm", se=FALSE)
#pass the ggplot to ggplotly
ggplotly(myplot)
```


===========================================================================
# Power analysis:

```{r}
library(pwr)
#compare powers
pwr.t.test(n=18, d=1, sig.level=.05,type='two.sample',alternative = 'two.sided')
pwr.t.test(n=18, d=1, sig.level=.05,type='paired',alternative = 'two.sided')
#compare sample size needed
pwr.t.test(power=.8, d=.5, sig.level=.05, type='two.sample',alternative = 'two.sided')
pwr.t.test(power=.8, d=.5, sig.level=.05, type='paired',alternative = 'two.sided')

# t-test:
# compute d (cohen's d) and ncp (non-centrality parameter)
tpower = read.table('example_input/tpower.dat',header=T, sep=" "); tpower
table = tpower%>% group_by(Condition) %>% summarize(mean = mean(Outcome),sd= sd(Outcome), n = n()); table
n1 = table$n[1]; n2 = table$n[2]; mean1 = table$mean[1]; mean2 = table$mean[2]
s1 = table$sd[1]; s2 = table$sd[2]; sp = sqrt((s1^2+s2^2)/2)
d = (mean2-mean1)/sp; d #effect size
delta = d*sqrt(50/2); delta #ncp (computed from d)
# compute power from alpha, df, ncp:
pt(q = qt(p=.975, df = 98), df=98, ncp=delta, lower=F) 
# compute power from d, n directly:
pwr.t.test(n=50,d=d, sig.level=.05,type='two.sample',alternative='two.sided')
#use ggplot for the null and alternative distributions with ncp:
data1 = tibble(value = rt(1000, df= 49, ncp= delta))
ggplot(data1, aes(value), xlim=c(-10,10)) + 
  stat_function(fun=dt,args=list(df=49, ncp=delta),col="dark blue",lwd=1.5,xlim=c(-10,10))+#curve for alt
  stat_function(fun=dt,args=list(df=49),col="dark red",lwd=1.5,xlim=c(-10,10))+#curve for null
  theme_classic() 

# f-test
#determine the power for a test on the covariate's effect 
fpower = read.table('example_input/fpower.dat',header=T, sep=" ")
null = lm(Outcome~Group, data=fpower)
full = lm(Outcome~Group+Covariate, data=fpower)
null.r2 = glance(reg0)$r.squared
full.r2 = glance(reg1)$r.squared
f2 = (full.r2-null.r2)/(1-full.r2)
ncp = f2*99
pf(q = qf(.95, 3, 95),df1=3,df2=95,ncp=ncp,lower=F)
pwr.f2.test(u=3, v=95, f2= f2, sig.level = .05)
```



===========================================================================
# Check model assumptions:

```{r}
#1) Checking for outlier
oly=read.table('example_input/olympics.dat',header=TRUE)
head(oly)
fit = lm(rating~comps+years,oly)
# i. leverage: hatvalue hii
oly$lvg = hatvalues(fit)
oly[1:5, c(2,3,4)]
# ii. discrepancy
rstudent(fit)
# iii. influence
dffits(fit)
cooks.distance(fit)
library(olsrr)
ols_plot_dffits(fit)
ols_plot_cooksd_chart(fit)
ols_plot_cooksd_bar(fit)
dfbetas(fit)
ols_plot_dfbetas(fit)
influence.measures(fit)
library(car)
influencePlot(fit)


# 2) Checking for multicollinearity
oly2=read.table('example_input/olympics.dat',header=TRUE)
head(oly2)
# i. correlation between the vars
cor(oly2[c(1,2,3)])
# ii. caution against Squared Multiple Correlation (SMC) >.6
library(psych)
smc(oly2[c(1,2,3)])
# iii. caution against tolerance (1-SMC) <.4
1-smc(oly2[c(1,2,3)])
# iv. caution against Variance Inflation Factor (VIF) >10
fit = lm(rating~years+comps,oly2)
vif(fit)


# 3) Checking for data normality 
rt = read.table("example_input/data_react.csv",header = TRUE,sep = ",") 
#This is to test if normality is significantly violated
shapiro.test(rt$RT)
#This is to plot the histogram with normal curve overlaid
library(ggplot2)
ggplot(rt, aes(RT)) + 
  geom_histogram(aes(y = ..density..), color='black', fill='gold', bins = 30) +
  stat_function(fun = dnorm, args = list(mean=mean(rt$RT), sd=sd(rt$RT)), col="dark red", lwd = 1.5) +
  theme_classic()
#This is to plot the Q-Q plot
ggplot(rt, aes(sample=RT)) + 
  stat_qq()+
  stat_qq_line(color="black")


# 4) Checking for residual normality 
RT = read_excel("example_input/data_rt.xlsx",1)
fit1 = lm(cfmt~benton_rt,data=RT)
fit2 = lm(cfmt~log(benton_rt),data=RT)
qqnorm(rstandard(fit1))
qqnorm(rstandard(fit2))
ols_plot_diagnostics(fit)
```

```{r}
RT = read_excel("example_input/data_rt.xlsx",1)
#get X and Y
Y = RT$cfmt
X = RT$benton_acc
# fit linear model & get b0, b1 for Yhat
lmfit = lm(Y~X)
b0 = tidy(lmfit)$estimate[1]
b1 = tidy(lmfit)$estimate[2]
# linear model for Yhat without error
Y_hat = b0 + b1*X
# plot the line of Y versus Yhat
lmfit2 = lm(Y~Y_hat)
p1 = gf_point(Y~Y_hat) %>%
  gf_smooth(method= 'lm',se=TRUE)
# check if the intercept approximates 0 and slope approx 1
intercept = tidy(lmfit2)$estimate[1]
slope = tidy(lmfit2)$estimate[2]
```



===========================================================================
Stats info 

```{r}
# descriptives
cd1 = read.table("example_input/data_classdata1.txt", header = TRUE)
Freq(cd1$TimeGrp) #check the frequency of each speed category
mean(cd1$Time)
mean(cd1$Time, trim = .10) #trim the top 10% and buttom 10% of the data before calculating mean
median(cd1$Time)
library(lsr)
modeOf(cd1$Time)
range(cd1$Time)
diff(range(cd1$Time))
quantile(cd1$Time)
IQR(cd1$Time)
var(cd1$Time)
sd(cd1$Time)
sqrt(var(cd1$Time))
describe(cd1$Time, IQR = TRUE)
describe(winsor(cd1$Time, trim = .10))

#generate random sample at specified size that ~ N(mean,sd) or U(min,max) or B(size,prob)
#normal
set.seed(100); rnorm(n=100, 16.5, 4.5) 
set.seed(1000); tibble(value = rnorm(1000, mean= 0, sd= 1))
set.seed(1000); tibble(value = rnorm(1000, mean= .967, sd= 1))
#uniform 
set.seed(100); runif(n=100,min=5,max=100) 
#binomial
set.seed(100); rbinom(n=100, size=10, prob=.5) #10 throws of fair coin, how many heads? do this 100 times

#binomial distribution: compute prob, plot binomial, find the p
library(gtools)
print('A(4,2) select 2 from 4: all possible selections with sequence considered')
nrow(permutations(n=4, r=2)) 
print('C(4,2) select 2 from 4: all possible selections, not considering sequence')
nrow(combinations(n=4, r=2)) 
#a basketball player makes 43% of his attempts. What is the precise probability that he will make 4 OR FEWER of his next 10 free throws? 
p.make4 = nrow(combinations(n=10,r=4))*((.43)^4)*((1-.43)^(10-4)); p.make3 = nrow(combinations(n=10,r=3))*((.43)^3)*((1-.43)^(10-3))
p.make2 = nrow(combinations(n=10,r=2))*((.43)^2)*((1-.43)^(10-2)); p.make1 = nrow(combinations(n=10,r=1))*((.43)^1)*((1-.43)^(10-1))
p.make0 = (1-.43)^10; p.fewer = p.make4 + p.make3 + p.make2 + p.make1 + p.make0; p.fewer
# or we can use the normal approximation of the binomial distribution
z = (4-(10*.43))/sqrt(10*.43*(1-.43)); pnorm(z,0,1,lower=F)
# generate binomial plot 
data = data.frame(x = 0:10, y = dbinom(0:10, size = 10, prob = .5))
ggplot(data, aes(x, y)) + geom_bar(stat = "identity", col = "blue", fill = "blue", width = .25) + labs(x = "Number of 'Heads'", y = "Probability") + theme_classic()
# find the prob/accumulative prob/quantile
print('P(X=25), the probability of getting 25 heads with 50 throws and a fair coin (p=.5)')
dbinom(x=25, size=50, prob=.5) 
print('P(X<=25), the cumulative probability of getting less or equal to 25 heads with 50 throws and a fair coin (p=.5)')
pbinom(q=25, size=50, prob=.5) 
print('x when P(X<=x)=.556, how many heads with 50 throws and a fair coin (p=.5) is the cumulative prob = .556?')
qbinom(p=.556,size=50,prob=.5)

#normal distirbution: find z-score, accumulative prob, prob density
print('calculate z-score from (sample.mean-pop.mean) / (pop.std / sqrt(sample.size))')
(1145-1060)/(199/sqrt(32))
print('find the z-score (from the accumulated-probability/area-under-the-curve)')
qnorm(0.9,mean=0,sd=1) #this returns the z-score of p=0.9 (the z-score that has the left area under curve of N(0,1) to be 0.9) 
qnorm(0.1,mean=0,sd=1) #this returns the z-score of p=0.1 (the z-score that has the left area under curve of N(0,1) to be 0.1) 
print('find the accumulated-probability/area-under-the-curve (from z-score)')
pnorm(1.9,mean=0,sd=1) #this returns the P(Z<1.9) (the area under curve left to z-score of 1.9) 
pnorm(-.9,mean=0,sd=1) #this returns the P(Z<-0.9) (the area under curve left to z-score of -0.9) 
pnorm(1145, mean=1060, sd=199/sqrt(32), lower=FALSE)
print('find the probability-density/y-of-point-on-curve (from z-score)')
dnorm(1.9,mean=0,sd=1) #this returns the p(Z=1.9) (the probability density, or, the value of the y-axis on the curve) 
dnorm(-.9,mean=0,sd=1) #this returns the p(Z=-.9) (the probability density, or, the value of the y-axis on the curve)

#t-distribution: find t/p
pt(q=0.2,df=8,lower=FALSE) #input t, get p; 
pt(q=0.2,df=18,lower=FALSE) #q=z-score, df --> shape of the t-distribution
pt(q=0.2,df=28,lower=FALSE)
qt(p=.025, df=8,lower=FALSE) #input p, get t
qt(p=.025, df=18,lower=FALSE) #p=alpha/2, df --> shape of the t-distribution
qt(p=.025, df=28,lower=FALSE)

#F-distribution: find f/p 
pf(q = .77, df1 = 1, df2 = 18, lower = FALSE) #input f, get p
qf(p = .77, df1 = 1, df2 = 18, lower = FALSE) #input p, get f

#Chi-sq distribution: find chi_sq/p  
pchisq(q = 13,df = 1,lower=F)
qchisq(p = .0003, df=1, lower=F)
```



===========================================================================
Stats tests 

t-test
```{r}
# 1)
# t-test
RT = read_excel("example_input/data_rt.xlsx",1)
TD= RT %>% filter(group=="0")
DP= RT %>% filter(group=="1")
resumes = read.csv("example_input/resumes.csv", header = TRUE)
pre = resumes$Pre
post = resumes$Post
#two independent samples t-test: comparing two vectors
t.test(TD$benton_acc,DP$benton_acc, alternative = "greater", var.equal = FALSE) 
t.test(TD$benton_rt,DP$benton_rt, alternative = "less", var.equal = FALSE)
#one sample t-test: comparing a vector with a population mu
t.test(TD$cfmt,mu=mean(na.omit(TD$cfmt))-2, alternative = "two.sided") 
#paired t-test
t.test(pre,post, alternative = "two.sided",paired=TRUE) 
```

Correlation test
```{r}
# 2)
# Correlation 
# correlation: see r table and plots
library(dplyr)
roi = read_excel("./example_input/data_roi.xlsx", sheet = 4)
roi_pre = roi %>% select(1:9,"chg_score")
scat_table = ggpairs(roi_pre,progress=F,lower=list(continuous=wrap("smooth",method="lm")))
corr_table = kable(cor(roi_pre, method = "pearson", use = "complete.obs"))
corr_table
# correlation test: get t/p/r of r
grades = read.table("./example_input/data_grades.dat", header = TRUE)
head(grades)
cor.test(x = grades$SAT, y = grades$GPA, alternative = "two.sided", method = "pearson", conf.level = .95)
#compare 2 correlations
library(DescTools); library(psych)
r1 = .2; r2= .6
z.r1 = FisherZ(rho = r1) #assume this is between X and Y
z.r2 = FisherZ(rho = r2) #assume this is between W and Z
paired.r(xy=r1, xz=r2, n=20, n2=18)
r.test(n=20,r12=r1,r34=r2,n2=18,twotailed=FALSE)
#r.test(n, r12, r34 = NULL, r23 = NULL, r13 = NULL, r14 = NULL, r24 = NULL, n2 = NULL,pooled=TRUE, twotailed = TRUE)
```

F-test(anova)
```{r}
# 3) 
# F-test (anova)
fit = lm(cfmt~ffmt, RT)
anova(fit)
library(broom)
tidy(anova(fit))
tidy(fit)
#coef
fit$coef
coef(fit)
#residual
fit$residuals
resid(fit)
#fitted
fit$fitted.values
fitted(fit)
#df residual
fit$df.residual
df.residual(fit)
#CI for b0 & b1
confint(fit)
```

One-way ANOVA
```{r}
# One-way between-subject ANOVA (1 fctr, >2 lvls)
waldo = read.table("example_input/waldo.dat", header = TRUE)
amodel = aov(search ~ cond, waldo)
#check model assumptions
plot(amodel) 
# model effect size
library(sjstats)
eta_sq(amodel)
omega_sq(amodel)
# visualize anova
ggplot(waldo, aes(cond, search)) +
  geom_bar(stat = "summary", fun.y = mean, aes(fill = cond)) +
  geom_errorbar(stat = "summary", fun.data = mean_se, width = .25) + theme_classic()
ggplot(waldo, aes(cond, search)) +
  geom_point(stat = "summary", fun.y = mean, aes(col = cond), size = 3) +
  geom_errorbar(stat = "summary", fun.data = mean_se, aes(col = cond), width = .25) +
  geom_line(lty = 2, stat = "summary", fun.y = mean, aes(group = 1)) + theme_classic()
#pairwise t test for further testing:
pairwise.t.test(x = waldo$search, g = waldo$cond, p.adj = "none")
pairwise.t.test(x = waldo$search, g = waldo$cond, p.adj = "bonferroni")
pairwise.t.test(x = waldo$search, g = waldo$cond, p.adj = "holm")
#planned linear regression
contrasts(waldo$cond) #check the default contrast
c1= c(-2, 1, 1) #"ctrl vs. load (info + time)"
c2= c(0, -1, 1) #"info vs. time"
contrasts(waldo$cond) = cbind(c1, c2) #overwrite the default contrast for planned-contrast
colnames(contrasts(waldo$cond)) = c("ctrl vs. load", "info vs. time")
contrasts(waldo$cond)
lmodel= lm(search ~ cond, waldo) 
summary(lmodel)$coefficients
# minimum significance threshold 
library(agricolae)
LSD.test(y = waldo$search, trt = waldo$cond, DFerror = 15, MSerror = 54.93, console = TRUE) #Fisher's LSD
#From table: qcrit <- 3.989; HSD <- qcrit * sqrt((MSw/n + MSw/n)/2); HSD
#groupdiffs <- c(xbar1-xbar2, xbar1-xbar3, xbar1-xbar4, xbar1-xbar5, xbar2-xbar3, xbar2-xbar4, xbar2-xbar5, xbar3-xbar4, xbar3-xbar5, xbar4-xbar5)
#abs(groupdiffs) > HSD; or, TukeyHSD(Ftest4)
HSD.test(y = waldo$search, trt = waldo$cond, DFerror = 15, MSerror = 54.93, console = TRUE) #Tukey's HSD
# simple ANOVA as linear model
tidy(lm(amodel))
```

Between-Subjects Factorial ANOVA
```{r}
#Factorial ANOVA (>1 fctr, >1 lvls per fctr)
quiz = read.table("example_input/quiz.dat", header = TRUE)
library(dplyr)
#the means table (cell means)
quiz %>%
  group_by(version, agegrp) %>%
  summarize(mean = mean(correct), n = n())
#ANOVA model
a = aov(correct ~ version + agegrp + version:agegrp, quiz)
a = aov(correct ~ version*agegrp, quiz)
Anova(a) #Type II SS 
Anova(a, type = "III") #Type III SS 
library(ggplot2)
#visualize: main effect of version 
ggplot(quiz, aes(version, correct)) +
  stat_summary(fun.y = mean, geom = "bar", aes(fill = agegrp), position = "dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", lwd = 1, width = .25, aes(group = agegrp), position = position_dodge(width = .9)) +
  labs(x = "Quiz Version", y = "Number Correct", fill = "Age Group") +
  theme_classic()
#visualize: main effect of age 
ggplot(quiz, aes(agegrp, correct)) +
  stat_summary(fun.y = mean, geom = "bar", aes(fill = version), position = "dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", lwd = 1, width = .25, aes(group = version), position = position_dodge(width = .9)) +
  labs(x = "Age Group", y = "Number Correct", fill = "Quiz Version") +
  theme_classic()
#visualize: interaction effect
ggplot(quiz, aes(agegrp, correct)) +
  stat_summary(fun.y = mean, geom = "point", aes(color = version), size = 5) +
  stat_summary(fun.y = mean, geom = "line", lwd = 2, aes(group = version, color = version)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(group = version, color = version), width = .25, lwd = 1) +
  labs(x = "Age Group", y = "Number Correct", fill = "Quiz Version") +
  theme_classic()
# Simple (conditional) effects
# compare between levels of a factor (across) at each of the levels of the other factor (fixed)
library(phia)
testInteractions(model = a, fixed = "version", across = "agegrp", adjust = "none") 
testInteractions(model = a, fixed = "agegrp", across = "version", adjust = "none")
# effect size
library(sjstats)
eta_sq(a)
eta_sq(a, partial = TRUE)
omega_sq(a)
omega_sq(a, partial = TRUE)
# contrasts and post-hoc tests
# CLD() outputs all pairwise marginal mean comparisons
library(emmeans)
library(multcompView)
CLD(emmeans(a, ~ version | agegrp), adjust = "none", details = TRUE)
CLD(emmeans(a, ~ agegrp | version), adjust = "none", details = TRUE)
# factorial ANOVA as linear model
lm(a)
```

Repeated-Measures ANOVA
```{r}
# Repeated-Measures ANOVA:
rmdata = read.table("example_input/rm.dat", header = TRUE)
library(tidyr); library(dplyr)
# convert wide to long format
rmlong = rmdata %>% gather(key = "cond", value = "search", -pID) 
rmlong$pID = factor(rmlong$pID)
#convert from long to wide: rmwide = spread(rmlong, cond, search)
# Error(pID/cond): cond is nested within participant (pID)
ws = aov(search ~ cond + Error(pID/cond), rmlong)
summary(ws)
# if Error NOT speficied, observations are assumed to be between-subject
bs = aov(search ~ cond, rmlong)
summary(bs)
# sphericity test 
# RM-ANOVA: independence between observations is violated, so need to see if variance of the difference between groups/conditions are the same 
ctrl_load = rmdata$Ctrl - rmdata$Load
ctrl_time = rmdata$Ctrl - rmdata$Time
load_time = rmdata$Load - rmdata$Time
c(var(ctrl_load), var(ctrl_time), var(load_time))
# ezANOVA: conduct ANOVA and auto-check sphericity, with 3 df output:
# df1 -- ANOVA table + ges (effect size measure); 
# df2 -- Mauchly's test for sphericity (not violated); 
# df3 -- Greenhouse Geisser (GGe) correction + Huynh-Feldt epsilon (HFe) correction
library(ez); ezANOVA(rmlong, dv = .(search), wid = .(pID), within = .(cond))
# effect size for RM:
library(sjstats)
eta_sq(rm); eta_sq(rm, partial = TRUE); 
omega_sq(rm); omega_sq(rm, partial = TRUE)
# post-hoc test for RM
pairwise.t.test(rmlong$search, rmlong$cond, paired = TRUE, p.adj = "bonferroni")
# visualize RM 
# exclude individual differences from errorbar (use adjusted score):
grandMean <- mean(c(rmdata$Ctrl, rmdata$Load, rmdata$Time))
rmdata$pMean <- (rmdata$Ctrl + rmdata$Load + rmdata$Time)/3
rmdata$adj <- grandMean - rmdata$pMean
rmdata$ctrl_adj <- rmdata$Ctrl + rmdata$adj
rmdata$load_adj <- rmdata$Load + rmdata$adj
rmdata$time_adj <- rmdata$Time + rmdata$adj
rmlong$search_adj <- c(rmdata$ctrl_adj, rmdata$load_adj, rmdata$time_adj)
rmlong %>% arrange(pID)
ggplot(rmlong, aes(x = cond, y = search_adj)) +
  stat_summary(fun = mean, geom = "bar", aes(fill = cond)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .25) +
  labs(x = "Search Condition", y = "Time to Find Waldo") +
  theme_classic()
```

Factorial RM ANOVA
```{r}
# Factorial RM ANOVA
recall = read.table("example_input/recall.dat", header = TRUE)
library(vtree); vtree(recall, vars = c("Subject", "Task", "Valence"))
#two-way-within model:
tww = aov(Recall ~ (Task * Valence) + Error(Subject/(Task * Valence)), recall)
#df(Residuals) in Subject:Task = df(Task) in Subject:Task x df(Residuals) in Subject 
summary(tww)
#means tables (2 ways to do so)
model.tables(tww, type = "means")
recall %>% group_by(Task, Valence) %>% summarise(n=n(),mean=mean(Recall))
# easy way to do the ANOVA model:
library(ez);ezANOVA(recall, dv = .(Recall), wid = .(Subject), within = .(Task, Valence), type = 3, detailed = TRUE)
```

Mixed ANOVA (within-between)
```{r}
# Mixed ANOVA
library(tidyr);library(dplyr); library(ez)
chat = read.table("example_input//chat.dat", header = TRUE); chat
chatlong = chat %>% gather(key = "Confed", value = "Convo", c(-ID, -Group))
chatlong %>% arrange(ID) #group x confed between-within subjects ANOVA
chatlong$ID = factor(chatlong$ID); chatlong$Confed = factor(chatlong$Confed)
#specify mixed model:
mixed = aov(Convo ~ Group*Confed + Error(ID/Confed) + Group, chatlong)
summary(mixed)
#ez mixed model: this output normally includes Mauchly's Test, but there is no sphericity concern here
#(because the within-subjects factor only has 2 levels)
ezANOVA(data = chatlong, dv = .(Convo), wid = .(ID), within = .(Confed), between = .(Group), type = 2, detailed = TRUE)
#adjust the values for the between-subjects error
tryMean <- mean(c(chat$Friendly[chat$Group=="Try"], chat$Unfriendly[chat$Group=="Try"])) # between-group mean1
norMean <- mean(c(chat$Friendly[chat$Group=="Normal"], chat$Unfriendly[chat$Group=="Normal"])) # between-group mean2
pMean <- (chat$Friendly + chat$Unfriendly)/2 #each participant's mean
#between-group factor mean - subject mean
adj <- NA; for(i in 1:nrow(chat)) {ifelse(chat$Group[i]=="Try",adj[i]<-tryMean-pMean[i],adj[i]<-norMean-pMean[i])}
chat$Friendly_adj <- chat$Friendly + adj; chat$Unfriendly_adj <- chat$Unfriendly + adj; chat
chatlong$Convo_adj <- c(chat$Friendly_adj, chat$Unfriendly_adj)
chatlong %>% group_by(Confed,Group) %>% summarize(n=n(),adjed = mean(Convo_adj), unadj= mean(Convo))
#visualize: unadjusted
library(ggplot2)
ggplot(chatlong, aes(Group, Convo)) +
  stat_summary(fun.y = mean, geom = "bar", aes(fill = Confed), position = "dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(group = Confed), width = .25, position = position_dodge(.90)) +
  theme_classic()
#visualize: adjusted 
ggplot(chatlong, aes(Group, Convo_adj)) +
  stat_summary(fun.y = mean, geom = "bar", aes(fill = Confed), position = "dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(group = Confed), width = .25, position = position_dodge(.90)) +
  theme_classic()
```

MANOVA (multipel DVs)
```{r}
# 4) 
# MANOVA
# short version (application)
move = read.table('example_input/movement.dat',header=T)
# using sleep category to predict for two dependent variables (speed and quick)
mm = manova(cbind(speed,quick) ~ sleep, move) #cbind: needs to be a matrix (can convert df to matrix using data.matrix)
#Using Pillai’s trace, we conclude that there was a significant effect of sleep conditions on speed and quickness, V = 0.46, F(4, 192) = 14.19, p < .001. 
summary(mm) 
#Separate univariate ANOVAs on the outcome variables revealed significant effects of sleep condition on both speed, F(2, 96) = 8.42, p < .001, and quickness, F(2, 96) = 23.78, p < .001.
summary.aov(mm) 
#DFA revealed 2 discriminant functions, accounting respectively for 81.4% and 18.6% of the variance. The first seems to be associated primarily with quickness, whereas the second is associated in opposing directions with speed. 
dfa = lda(sleep ~ speed + quick, move); dfa 
#Plots of these discriminant functions suggest that scores on LD1 (concerned primarily with quick) distinguish those in 3-hour sleep condition from those in the 9-hour sleep condition, whereas scores on LD2 (concerned primarily with speed) distinguish those in the 6-hour sleep condition from those in the other 2 conditions.
move$LD1 <- predict(dfa)$x[,1]
move$LD2 <- predict(dfa)$x[,2]
ggplot(move, aes(LD1, LD2)) +
  geom_text(size = 4, aes(label = sleep, color = sleep)) +
  theme_classic()
```
```{r}
# full version (computations)
library(mvoutlier)
words = read.table("example_input/words.dat", header = TRUE);words
aq.plot(words[, 3:4]) #adjusted quantitile plot for outliers based on our two DVs: column 3 and 4
#T
SSTnum <- sum((words$number-mean(words$number))^2)
SSTconf <- sum((words$confidence-mean(words$confidence))^2)
CPT <- sum((words$number-mean(words$number))*(words$confidence-mean(words$confidence)))
SSTnum; SSTconf; CPT
#H
SSMnum <- sum(33*(mean(words$number[words$speed == "fast"])-mean(words$number))^2,
    33*(mean(words$number[words$speed == "normal"])-mean(words$number))^2,
    33*(mean(words$number[words$speed == "slow"])-mean(words$number))^2)
SSMconf <- sum(33*(mean(words$confidence[words$speed == "fast"])-mean(words$confidence))^2,
    33*(mean(words$confidence[words$speed == "normal"])-mean(words$confidence))^2,
    33*(mean(words$confidence[words$speed == "slow"])-mean(words$confidence))^2)
CPM <- sum(33*((mean(words$number[words$speed == "fast"])-mean(words$number))*(mean(words$confidence[words$speed == "fast"])-mean(words$confidence))),
    33*((mean(words$number[words$speed == "normal"])-mean(words$number))*(mean(words$confidence[words$speed == "normal"])-mean(words$confidence))),
    33*((mean(words$number[words$speed == "slow"])-mean(words$number))*(mean(words$confidence[words$speed == "slow"])-mean(words$confidence))))
SSMnum; SSMconf; CPM
#E
SSTnum-SSMnum
SSTconf-SSMconf
CPT-CPM
#get test stats
T_mat <- matrix(c(2248.51, 45.57, 45.57, 745.00), nrow = 2)
H_mat <- matrix(c(621.90, 10.73, 10.73, 125.33), nrow = 2)
E_mat <- matrix(c(1626.61, 34.84, 34.84, 619.68), nrow = 2)
E_inv = solve(E_mat)
T_mat;H_mat;E_mat;E_inv
#get the initial $HE^{-1}$ matrix:
HEinv <- H_mat %*% E_inv; HEinv
#above can be obtained by extracting eigenvector coefficients:
mm <- manova(cbind(number, confidence) ~ speed, words)
library(candisc); candisc(mm)$coeffs.raw
#check data dimensions
# eyeball it:
library(ggplot2)
ggplot(words, aes(number, confidence)) +
  geom_point() +
  scale_x_continuous(limits = c(5,35)) +
  scale_y_continuous(limits = c(-5,15))
# draw boundry:
find_hull <- function(words) words[chull(words$number, words$confidence), ]
hulls <- find_hull(words)
ggplot(words, aes(number, confidence)) +
  geom_point() +
  geom_polygon(data = hulls, alpha = .2)
# circle it:
library(ggforce);
ggplot(words, aes(number, confidence)) +
  geom_point() +
  geom_ellipse(aes(x0=mean(number)+1.5,y0=mean(confidence),a=12,b=8,angle=pi/10))
#Using the coefficients to compute discriminant function variates
words$V1 <- candisc(mm)$coeffs.raw[1,1]*words$number + candisc(mm)$coeffs.raw[2,1]*words$confidence
words$V2 <- candisc(mm)$coeffs.raw[1,2]*words$number + candisc(mm)$coeffs.raw[2,2]*words$confidence
#Creating $HE^{-1}$ matrix from the discriminant function variates:
MV <- matrix(c(36.71, 0, 0, 19.41), nrow = 2)
EV <- matrix(c(96, 0, 0, 96), nrow = 2)
HEVinv <- MV %*% solve(EV);
#The off-diagonal values of the new $HE^{-1}$ matrix (cross-products) are 0
#The diagonal values are the *eigenvalues* of the original $HE^{-1}$
HEVinv
#or, we can also obtain the eigenvectors from the initial HEinv:
eigen(HEinv)$values
# compare test statistics to an F-distribution
summary(mm, test = "Pillai")
summary(mm, test = "Hotelling-Lawley")
summary(mm, test = "Wilks")
summary(mm, test = "Roy")
# visualizing MANOVA results
wordslong <- words %>% gather(key = "DV", value = "value", c(-pID, -speed, -V1, -V2)); wordslong %>% arrange(pID)
ggplot(wordslong, aes(speed, value)) +
  stat_summary(fun = mean, geom = "bar", aes(fill = DV), position = "dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(group = DV),position=position_dodge(.90),lwd = 1,width = .25) +
  theme_classic()
# post hoc univariate tests: linear discriminant analysis
summary.aov(mm)
library(MASS)
DA <- lda(speed ~ number + confidence, words)
#V1 appears to be affected by number but not by confidence, 
#V2 appears to be affected by confidence but not by number
#The "trace proportions": ratios of each eigenvalue (corresponding to V1,V2) to the sum of the eigenvalues
DA
head(predict(DA)$x)
#visualize: the values for each variate that we calculated as V1 and V2
slope1 <- -0.243033440/0.005648851
slope2 <- 0.004951098/-0.393795017
plot(DA, col = c("red", "blue", "green4"))
abline(a = 0, b = slope1)
abline(a = 0, b = slope2)
plot(DA)
#prettier visualize:
words$V1_cent <- words$V1 - mean(words$V1)
words$V2_cent <- words$V2 - mean(words$V2)
ggplot(words, aes(-V1_cent, -V2_cent)) +
  geom_text(size = 4, aes(label = speed, color = speed)) +
  geom_abline(intercept = 0, slope = slope1, lwd = 1) +
  geom_abline(intercept = 0, slope = slope2, lwd = 1) +
  labs(x = "Centered V1", y = "Centered V2") +
  scale_x_continuous(limits = c(-4, 4)) +
  scale_y_continuous(limits = c(-3, 3)) +
  theme_classic()
#Values for V1 (mostly affected by number) appear to distinguish those in the slow condition from those in the fast 
#Values for V2 (mostly affected by confidence) appear to distinguish those in the normal from those in the other 2 conds
```

ANCOVA (analysis of covariance)
```{r}
library(ggplot2)
waldo = read.table('example_input/waldo2.dat',header=T, sep=' '); waldo
anova = aov(search~cond,waldo); summary(anova) #normal anova
ancova = aov(search~cond+exp,waldo); summary(ancova) #exp-erience is a covaraite (continuous)
anova(anova,ancova) #nested model F-test when added var=1 == t-test for the added predictor (f=t^2)

# assumption check
# 1) homogeneity of regression slopes: not violated
ggplot(waldo,aes(x=exp, y=search))+ geom_point(aes(col=cond)) + 
  geom_smooth(method='lm',se=F,aes(col=cond),fullrange=T) #plot
slopecheck = aov(search~exp*cond,waldo); Anova(slopecheck) #test
# 2) predictor-covariate independence: not violated
pci = aov(exp~cond,waldo); summary(pci)

#adjusted means and plot
lm1 = lm(search~exp+cond,waldo); tidy(lm1) #a linear model for ANCOVA
waldo$search_adj = waldo$search - (tidy(lm1)$estimate[2])*(waldo$exp - mean(waldo$exp))
ggplot(waldo, aes(x=cond,y=search_adj))+ stat_summary(fun.y=mean, geom='bar', aes(fill=cond))+
  stat_summary(fun.data=mean_se, geom='errorbar', width=.25)

#center covariate to make intercept more interpretable
waldo$c_exp = waldo$exp - mean(waldo$exp)
lm2 = lm(search~ c_exp+ cond, waldo); summary(lm2) # #a linear model for ANCOVA
```


Binomial-test
```{r}
# 5) 
# Binomial test/ sign-test
# Consider a coin flip (p=.5): what's the chance of getting 75 heads with 100 throws?
binom.test(x = 55, n = 100, p = .50) 
# With 14 improvement out of 20 students, what's the evidence for improvement? 
midterm <- c(76,70,64,77,80,84,70,89,68,61,76,76,74,96,80,78,78,69,73,80)
final <- c(78,82,65,79,82,88,90,92,69,70,79,78,75,99,77,72,64,62,70,79)
library(coin)
sign_test(midterm ~ final) #test for change of sign (pos, neg) with pre to post
pnorm(q=14, mean=20*.5,sd=sqrt(20*.5*.5), lower=F)*2 #normal approximation: 14 improved out of 20 in the norm-dist
binom.test(x = 14, n = 20, p = .50) #binomial: how likely 14 out of 20 improved with .5 being chance level
```

Chi-Squared-test
```{r}
# 6)
#Chi_sq test of goodness-of-fit   
library(coin)
# assume three drink flavors were equally popular; we sold 120 with the below observed frequency for each (37,58,25) 
chocolate <- 37
vanilla <- 58
mint.chip <- 25
ice.cream <- cbind(chocolate, vanilla, mint.chip)
rownames(ice.cream) <- "Observed"
colnames(ice.cream) <- c("Chocolate", "Vanilla", "Mint Chip")
#Under the null hypothesis, the frequencies of each of the 3 flavors sold should all equal to 40. As test statistics are computed and confirmed below, $\chi^2$(2) = 13.95, p<.001. Therefore, we reject the null hypothesis and conclude that the popularity of the three flavors is different from what would expected if they were equally popular.
chisq.test(x = ice.cream, p = rep(1/3,3))

#Chi_sq test of independence
no.pass <- c(39, 74)
one.pass <- c(28, 40)
two.or.more.pass <- c(18, 51)
run_red <- cbind(no.pass, one.pass, two.or.more.pass)
rownames(run_red) <- c("Ran Light", "Did Not Run Light")
colnames(run_red) <- c("No Passengers", "One Passenger", "2+ Passengers")
addmargins(run_red) 
# H0: The distributions of the 2 variables (i.e. run through red lights or not, and #of passenger on the car) are independent. As test statistics computed and confirmed below, $\chi^2$(2) = 3.50, p=.174. Therefore, we fail to reject the null hypothesis, and conclude that the distribution of these two variables are independent.
chisq.test(x = run_red, correct = FALSE)
```



===========================================================================
# Effects (moderation/mediation):

Moderation (interaction) effect:
```{r}
library(dplyr)
library(pequod)
library(interactions)

mag = read.table('example_input/mag.dat',header=T)
mag = mag %>% mutate(c_price=price-mean(price),c_stocked=stocked-mean(stocked)); #center predictors
m1 = lm(sales~c_price+c_stocked,mag);summary(m1)
m2 = lm(sales~c_price*c_stocked,mag); summary(m2)
# simple slope 
  # check predictor's effect at -1/+1 SD of the moderator)
  # 1) R command 'simpleSlope'
  library(pequod)
  m3 = lmres(sales~price*stocked,centered=c('stocked','price'),mag)
  ss = simpleSlope(m3,pred='stocked',mod1='price'); ss #predictor being stocked, moderator being price 
  # 2) calculate:
  mag = mag %>% mutate(lo_price = c_price + sd(c_price), hi_price = c_price - sd(c_price))
  m4 = lm(sales~lo_price*c_stocked,mag); summary(m4)$coefficients
  m5 = lm(sales~hi_price*c_stocked,mag); summary(m5)$coefficients
# johnson neyman: 
  # get the range of the moderator within which the predictor's effect is significant
  library(interactions)
  johnson_neyman(model=m2, pred='c_stocked', modx='c_price')
# plot the moderation 
  #at -1/+1 SD of the moderation
    PlotSlope(ss) 
  #at -1/0/+1 SD of the moderation
    interact_plot(model=m2,pred='c_stocked',modx='c_price',line.thickness = 3)

psych = read.table('example_input/exams.dat',header=T)
psych = psych %>% mutate(c_career=career-mean(career)); psych #center type  
# Hierachical regression for moderation
  #public as reference
      contrasts(psych$type) = c(1,0); contrasts(psych$type) 
      #first step
      m1 = lm(score~type+c_career,psych); summary(m1)
      #second step
      m2 = lm(score~type*c_career,psych); summary(m2)
  #private as reference
      contrasts(psych$type) = c(0,1); contrasts(psych$type) 
      #new model with interaction:
      m3 = lm(score~type*c_career,psych)
#interaction plot & simple slopes
  #with 1 predictor 1 moderator
      psych$type = as.numeric(psych$type)
      lmc3 = lm(score~type*career,psych)
      interact_plot(model=lmc3,pred='career',modx='type',line.thickness = 3)
      sim_slopes(model=lmc3,pred='career',modx='type')
  #with 1 predictor 2 moderators
      tw = lm(score~type*family*c_career,psych); summary(tw)
      interact_plot(model=tw, pred='c_career', modx='type', mod2='family')
      sim_slopes(model=tw,pred='c_career', modx='type', mod2='family')
#controlling for the covariate:  
    nm = lm(score~type*family+c_career,psych);summary(nm); anova(nm)
    interact_plot(model=nm, pred='type',modx='family', mod2='c_career')
```

Mediation effect (a predicts c via b)
```{r}
library(broom)
wtest = read.table('example_input/wtest.dat',header=T)
#path c (predictor-outcome):
total = lm(score ~ sleep, wtest); tidy(total)
#path a (predictor-mediator): 
path_a = lm(WPM ~ sleep, wtest); tidy(path_a)
#path b (coef of mediator controlled for predictor) 
path_b = lm(score ~ WPM + sleep, wtest); tidy(path_b)$estimate[2]
#path c' -- direct effect (coef of predic/tor controlled for mediator): 
tidy(path_b)$estimate[3]
#indirect effect
  #1) a_hat*b_hat:
  a_hat = as.numeric(coef(path_a)[2]); b_hat = as.numeric(coef(path_b)[2])
  IE = a_hat * b_hat; IE
  #2) total effect - direct effect
  coef(total)[2]-coef(path_b)[3]
#Sobel test (for the indirect effect, i.e. mediator effect):
  #compute
  SE_a = summary(path_a)$coefficients[2,2]; SE_b = summary(path_b)$coefficients[2,2]
  sIE = sqrt((a_hat^2 * SE_b^2) + (b_hat^2 * SE_a^2))
  SobelZ = IE/sIE; SobelZ; pnorm(SobelZ, lower=F)*2
  #check 
  library(multilevel)
  sobel(pred=wtest$sleep, med=wtest$WPM, out=wtest$score)

#bootstrapped mediation analysis
library(boot)
  #returns coef of predictor and mediator
  bootReg <- function(formula, data, i) { 
    fit <- lm(formula, data = data[i,]) 
    return(coef(fit)[2:3])} 
  #returns indirect effect
  bootMed <- function(formula1, formula2, data, i) {
    fit1 <- lm(formula1, data = data[i,])
    fit2 <- lm(formula2, data = data[i,]) 
    return(coef(fit1)[2]-coef(fit2)[2])} 
#bca CI for the direct effect
set.seed(1)
bootResults = boot(data = wtest, formula = lm(score ~ WPM + sleep, wtest), statistic = bootReg, R=2000)
bootResults$t0[1]; bootResults$t[1:5,1]
boot.ci(bootResults, type='bca', index=1)
boot.ci(bootResults, type='bca', index=2)
#bca CI for the indirect effect
set.seed(1)
bootMedResults = boot(data=wtest,statistic=bootMed,
                      formula1=lm(score~sleep, wtest),
                      formula2=lm(score~sleep+WPM, wtest), R=2000)
bootMedResults$t0; bootMedResults$t[1:5,]; hist(bootMedResults$t)
boot.ci(bootMedResults, type='bca')
#alternative command
library(mediation)
med = mediate(model.m=path_a, model.y=path_b, treat='sleep',mediator='WPM', 
              sims=2000, boot=T, boot.ci.type='bca'); summary(med)

```


===========================================================================
# Fit regression model:

```{r}
library(tidyverse)

# 1) simple/multiple linear regression
fit0 = lm(cfmt~1,data=RT)
fit1 = lm(cfmt~benton_acc,data=RT)
fit2 = lm(cfmt~benton_rt1,data=RT)
fit3 = lm(cfmt~benton_acc+benton_rt1,data=RT)
library(broom)
tidy(fit1)
glance(fit1)
summary(fit1)
kable(anova(fit1))
# get correlation coefficients
library(ppcor)
oly=read.table('example_input/olympics.dat',header=TRUE)
head(oly)
tidy(lm(rating~comps+years,oly))
#semi-partial correlation coefficients
spcor(oly)
spcor(oly)$estimate[3, 1] #this is spcorr for comps~rating
#partial correlation coefficients
pcor(oly)
pcor(oly)$estimate[3, 1] #this is pcorr for comps~rating


# 2) binomial logistic regression
# example 1:
  RT = read_excel("example_input/data_rt.xlsx",1)
  nullmod = glm(group~1, data=RT, family='binomial')
  fit4 = glm(group~benton_acc, data=RT, family='binomial'); tidy(fit4); glance(fit4)
  pR2=1-logLik(fit1)/logLik(nullmod);pR2 # McFadden's Pseudo-R2
# example 2:
  # Interpret coefficients:
  tennis = read.table('example_input/tennis.dat',header=T)
  fit1 = glm(outcome~practice, data=tennis,family = binomial(link='logit')); summary(fit1)
    # 1) intercept: log odds of winning, i.e. logit(win), is -2.19 at 0-hour of practice; 
                    #odds of winning is .11 with no practice; 
                    #the overall probability of winning is 10% without practice.
      coef(fit1)[1]; exp(coef(fit1)[1]); phat = (0.1121717)/(1+0.1121717); phat
    # 2) predictor: Logit(win) increased by 24% for every 1-unit increase in practice; 
                    #odds of winning is multiplied by 1.28 (increased 28%): 
                                #for every 1-unit increase in practice time.  
      coef(fit1)[2]; exp(coef(fit1)[2])
  # Plot the model:
    library(ggplot2); library(dplyr); 
    tennis =tennis%>%mutate(win=1*(outcome=="win"))
    ggplot(data=tennis, aes(x=practice, y=win))+ geom_point(size=2, color='dark blue')+
      geom_smooth(se=F, method='glm', method.args=list(family=binomial), lwd=1.5,color='dark red')
  # Include a second predictor:
  fit2 = glm(outcome~practice+workout, data=tennis,family = binomial(link='logit')); summary(fit2)
    # 1) intercept: The person's log odds of winning is 4.12, 
                  #odds of winning is 61.56 with no practice or workout.
      coef(fit2)[1]; exp(coef(fit2))[1]
    # 2) predictor1: Logit(win) increases by .23 for every 1-unit increase in practice time; 
                    #odds of winning is multiplied by 1.26 (increased by 26%) 
                                        #for every 1-unit increase in practice.
      coef(fit2)[2]; exp(coef(fit2))[2]
    # 3) predictor2: Logit(win) decreases by .36 for every 1-unit increase in workout time; 
                    #odds of winning is multiplied by .70 (decreased by 30%)        
                                        #for every 1-unit increase in workout.
      coef(fit2)[3]; exp(coef(fit2))[3]
  # Likelihood Ratio test (lr test):
  lr = deviance(fit1)-deviance(fit2)
  lr; pchisq(q=lr, df=1, lower=F)
  library(lmtest)
  lrtest(fit2, fit1)

# 3) multinomial logistic regression
petdata = read.table('example_input/petowner.dat',header=T); summary(petdata)
library(mlogit)
petwide = mlogit.data(petdata, choice='pet', shape='wide')
mlr = mlogit(pet~1 | extraversion, data=petwide, reflevel = 'fish'); summary(mlr)
    #Log odds of pet = bird (compared to pet = fish) is 0.72770; the odds of pet = bird is multiplied by 2.070308 for every 1-unit increase in extraversion. 
    #Log odds of pet = cat (compared to pet = fish) is 0.71042; the odds of pet = cat is multiplied by 2.034844 for every 1-unit increase in extraversion.  
    #Log odds of pet = dog (compared to pet = fish) is 0.50864; the odds of pet = bird is multiplied by 1.663028 for every 1-unit increase in extraversion.  
    #Log odds of pet = rabbit (compared to pet = fish) is 0.89053; the odds of pet = bird is multiplied by 2.436430 for every 1-unit increase in extraversion. 
exp(mlr$coefficients)[5:8]    

# 4) robust regression
library(MASS)
library(sfsmisc)
fit5 = rlm(com~benton_acc+benton_rt1, data=TD) 
tidy(fit5)
f.robftest(fit5,var='benton_rt1')
library(robust)
fit6 = lmRob(com~benton_acc+benton_rt1, data=TD) 
fit7 = glmRob(group~benton_acc+benton_rt1, data=RT, family = "binomial") 
tidy(fit6)
tidy(fit7)

# 5) mixed effects regression
library("lmerTest")
roi = read_excel("/Users/lixianolivia/Desktop/PrcpLearn-fMRI2/furtherAnalysis/2_ROI/roi_data_clean.xlsx", sheet = 1)
fit = lmer(score~ (1|sub) + PO_l + time+ time*PO_l, data=roi) #rmi=random intercept model
summary(fit)
glance(fit)
```



===========================================================================
# Compare model fit:

```{r}
#anova test for nested models
test1 = anova(fit1,fit3)
test2 = anova(fit0,fit1)
test3 = anova(nullmod,fit4)

#likelihood ratio test for nested models
test3 = lrtest(fit0,fit1)
test4 = lrtest(fit1,fit3)
test5 = lrtest(fit2,fit3)
test6 = lrtest(nullmod,fit4)

# forward stepwise
library(orcutt)
library(olsrr)
fit=lm(com~benton_acc+benton_rt1+benton_rt2+age+gender+race,data=TD)
forward.sequence = ols_step_forward_p(fit, penter = 0.20)
plot(forward.sequence)
ols_step_forward_aic(fit)
vif(fit) # check if heavily collineared (>10) 

# backward stepwise
backward.sequence = ols_step_backward_p(fit, penter = 0.20)
plot(backward.sequence)
ols_step_backward_aic(fit)

# bidirectional stepwise
bidirect.sequence = ols_step_both_p(fit, penter = 0.20)
plot(bidirect.sequence)
ols_step_both_aic(fit)

# stepwise alternatives 
library(MASS)
mydata = na.omit(TD[3:14])
fit_full = lm(com~benton_acc+benton_rt1+benton_rt2+age+gender+race,data=mydata)
fit_init = lm(com ~ 1, data=mydata)
bw = stepAIC(fit,direction="backward")
fw = stepAIC(fit_init,direction="forward",scope=list(upper=fit_full))
bd = stepAIC(fit_init,direction="both",scope=list(upper=fit_full,lower=fit_init))

```



```{r}
library(mosaic)
library(modelr)
library(lmtest) #all sorts of assumption tests for linear model, e.g. dwtest

```





=====================================ROC curve===========================================

1. Generate the signal and noise Guassian distributions according to FA and Hit  
2. Create the ROC curve (plotting the noise distribution against the signal distribution)

```{r}
#assume this is our FA nad Hit:
FA = .3
Hit = .45

#calculate the d and beta
c = qnorm(FA,lower=FALSE)
cs = qnorm(Hit,lower=FALSE)
d = c-cs
fs.c = dnorm(cs)
fn.c = dnorm(c)
beta = fs.c/fn.c

#creating the noise and signal+noise curves: 
x = seq(-5,5,by=0.1) # setting the range of signal
PX_n = pnorm(x, mean = 0, sd = 1,lower=FALSE) # P(X > c | noise only) = False alarm rate
PX_s = pnorm(x, mean = d, sd = 1, lower=FALSE) # P(X > c | signal plus noise) = Hit rate
df = data.frame(x, PX_n, PX_s)

#plotting the ROC curve with our beta criterion marked out (slope of tangent line at the blue dot) 
roc = ggplot(df) + xlab("P('yes'| noise)") + ylab("P('yes'| signal)") +
  geom_line(aes(PX_n, PX_s)) + 
  geom_point(aes(x=FA, y=Hit), colour="blue",size=5)+
  geom_abline(slope=beta, intercept=Hit-FA*beta, color='blue',lty='twodash')+
  geom_abline(slope = 1) +
  ggtitle("ROC Curve") + 
  coord_equal()
roc
```

Still use the last example:
1. Assume that we have 100 trials per condition (noise 100 trials, signal 100 trials). 
   FA=.3 (100 noise: 30 'yes', 70 'no')
   Hit=.45 (100 signal: 45 'yes', 55 'no')
2. Contruct the roc curve just from these numbers:

```{r}
library(pROC)
#noise trials: 'no'-70, 'yes'-30
control =  c(rep(0, 70), rep(1, 30)) 
#signal trials: 'no'-55, 'yes'-45
case = c(rep(0, 55), rep(1, 45)) 
#specificity: inverse of FA
sdt = roc(controls=control, cases=case)
plot(sdt)
```


```{r}
library(pROC)
#compute ROC curve from data; plot it
data(aSAH)
aSAH
roc.s100b <- roc(aSAH$outcome, aSAH$s100b)
roc.wfns <- roc(aSAH$outcome, aSAH$wfns) 
roc.ndka <- roc(aSAH$outcome, aSAH$wfns)
plot(roc.s100b)

# Add a smoothed ROC:
plot(smooth(roc.s100b), add=TRUE, col="blue") 
legend("bottomright", legend=c("Empirical", "Smoothed"), col=c(par("fg"), "blue"), lwd=2)

# Plot the AUC 
plot(roc.s100b, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2), grid.col=c("green", "red"), max.auc.polygon=TRUE, auc.polygon.col="lightblue", print.thres=TRUE)

# To plot a different partial AUC, we need to ignore the existing value # with reuse.auc=FALSE:
plot(roc.s100b, print.auc=TRUE, auc.polygon=TRUE, partial.auc=c(1, 0.8),
partial.auc.focus="se", grid=c(0.1, 0.2), grid.col=c("green", "red"), max.auc.polygon=TRUE, auc.polygon.col="lightblue",
print.thres=TRUE, print.thres.adj = c(1, -1),
reuse.auc=FALSE)
# Add a second ROC curve to the previous plot: 
plot(roc.wfns, add=TRUE)

# Plot some thresholds, add them to the same plot
plot(roc.ndka, print.thres="best", 
     print.thres.best.method="youden")
plot(roc.ndka, print.thres="best",
     print.thres.best.method="closest.topleft", add = TRUE)
plot(roc.ndka, print.thres="best",
     print.thres.best.method="youden", 
     print.thres.best.weights=c(50, 0.2), 
     print.thres.adj = c(1.1, 1.25), add = TRUE)

#creating the noise and signal+noise curves: 
x = seq(-5,5,by=0.1) # setting the range of signal
PX_n = pnorm(x, mean = 0, sd = 1,lower=FALSE) # P(X > c | noise only) = False alarm rate
PX_s = pnorm(x, mean = 0.3987392, sd = 1, lower=FALSE) # P(X > c | signal plus noise) = Hit rate
df = data.frame(x, PX_n, PX_s)
df
```


=====================================network diagram===========================================

```{r}
library(igraph)
library(ggraph)
```

```{r}
# Load researcher data
dataUU <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/13_AdjacencyUndirectedUnweighted.csv", header=TRUE)

# Transform the adjacency matrix in a long format
connect <- dataUU %>% 
  gather(key="to", value="value", -1) %>%
  na.omit()

# Number of connection per person
c( as.character(connect$from), as.character(connect$to)) %>%
  tibble::enframe(name = NULL) %>%
  group_by(value) %>%
  summarise(n=n()) -> coauth
colnames(coauth) <- c("name", "n")

# Create a graph object with igraph
mygraph <- graph_from_data_frame( connect, vertices = coauth )

# Make the graph
ggraph(mygraph, layout="fr") + 
  #geom_edge_density(edge_fill="#69b3a2") +
  geom_edge_link(edge_colour="black", edge_alpha=0.2, edge_width=0.3) +
  geom_node_point(aes(size=n, alpha=n)) +
  theme_void() +
  theme(
    legend.position="none",
    plot.margin=unit(rep(1,4), "cm")
  ) 
```

```{r}
# Make the graph
ggraph(mygraph, layout="drl") + 
  #geom_edge_density(edge_fill="#69b3a2") +
  geom_edge_link(edge_colour="black", edge_alpha=0.2, edge_width=0.3) +
  geom_node_point(aes(size=n, alpha=n)) +
  theme_void() +
  theme(
    legend.position="none",
    plot.margin=unit(rep(1,4), "cm")
  ) 
```

```{r}
ggraph(mygraph, layout="igraph", algorithm="randomly") + 
  #geom_edge_density(edge_fill="#69b3a2") +
  geom_edge_link(edge_colour="black", edge_alpha=0.2, edge_width=0.3) +
  geom_node_point(aes(size=n, alpha=n)) +
  theme_void() +
  theme(
    legend.position="none",
    plot.margin=unit(rep(1,4), "cm")
  ) 
```

```{r}
# Libraries
library(igraph)
library(networkD3)

# create a dataset:
data <- data.frame(
  from=c(rep(1,9),rep(2,9),rep(3,9),rep(4,9),rep(5,9),rep(6,9),rep(7,9),rep(8,9),rep(9,9)),
  to=c(rep(c(1:9),6), rep(1:3,9)) )

# Plot
p <- simpleNetwork(data, height="100px", width="100px")
p

# save the widget
# library(htmlwidgets)
# saveWidget(p, file=paste0( getwd(), "/HtmlWidget/networkInteractive1.html"))
```

```{r}
# Libraries
library(igraph)
library(networkD3)

# create a dataset:
data <- data_frame(
  from=c("A", "A", "B", "D", "C", "D", "E", "B", "C", "D", "K", "A", "M"),
  to=c("B", "E", "F", "A", "C", "A", "B", "Z", "A", "C", "A", "B", "K") )

# Plot
p <- simpleNetwork(data, height="100px", width="100px",        
        Source = 1,                 # column number of source
        Target = 2,                 # column number of target
        linkDistance = 10,          # distance between node. Increase this value to have more space between nodes
        charge = -900,                # numeric value indicating either the strength of the node repulsion (negative value) or attraction (positive value)
        fontSize = 14,               # size of the node names
        fontFamily = "serif",       # font og node names
        linkColour = "#666",        # colour of edges, MUST be a common colour for the whole graph
        nodeColour = "#69b3a2",     # colour of nodes, MUST be a common colour for the whole graph
        opacity = 0.9,              # opacity of nodes. 0=transparent. 1=no transparency
        zoom = T                    # Can you zoom on the figure?
        )
p

# save the widget
# library(htmlwidgets)
# saveWidget(p, file=paste0( getwd(), "/HtmlWidget/networkInteractive2.html"))
```


=====================================scripts wrote earlier===========================================

```{r}
library(readxl) 
library(dplyr)
library(psych)
library(DescTools)

filelist = list.files('/Users/xianl/Desktop/FreeR_data/results/plots/try/','.xlsx')
print(filelist)
counter=0
sub.merge=tibble()
corr = tibble()
for(i in 1:length(filelist)) {
  for (j in 1:30){
    try.name = paste('sub',as.character(j),'.xlsx',sep='')
    if (grepl(try.name,filelist[i])==TRUE){
      print(try.name)
      
      full.path = paste('/Users/xianl/Desktop/FreeR_data/results/plots/try/',try.name,sep='')
      sub = read_excel(full.path,1)
      
      name = paste('sub',as.character(j),sep='')
      caus = as.numeric(sub$CausalityCent)
      caus_z = (caus-mean(caus))/sd(caus)
      sem = as.numeric(sub$SemanticCent)
      sem_z = (sem-mean(sem))/sd(sem)
      rcl = as.numeric(sub$RecallProb)
      rcl_z = (rcl-mean(rcl))/sd(rcl)
      
      sub = tibble(sub.num=j, 
         event=as.numeric(sub$Events),
         sem.z = sem_z,
         caus.z= caus_z, 
         rcl.z = rcl_z,
         recall=sub$RecallProb, 
         sem=sub$SemanticCent, 
         caus=sub$CausalityCent)
      sub.merge=rbind(sub.merge,sub)
      
      
      sem.rcl = cor.test(sem_z,rcl_z)
      caus.rcl = cor.test(caus_z,rcl_z)
      sem.caus = cor.test(sem_z,caus_z)
      sub.corr = tibble(sub.num=j,
        sem.rcl.r = tidy(sem.rcl)$estimate,
        sem.rcl.p = tidy(sem.rcl)$p.value,
        sem.rcl.95ci.low = tidy(sem.rcl)$conf.low,
        sem.rcl.95ci.high = tidy(sem.rcl)$conf.high,
        
        caus.rcl.r = tidy(caus.rcl)$estimate,
        caus.rcl.p = tidy(caus.rcl)$p.value,
        caus.rcl.95ci.low = tidy(caus.rcl)$conf.low,
        caus.rcl.95ci.high = tidy(caus.rcl)$conf.high,
        
        sem.caus.r = tidy(sem.caus)$estimate,
        sem.caus.p = tidy(sem.caus)$p.value,
        sem.caus.95ci.low = tidy(sem.caus)$conf.low,
        sem.caus.95ci.high = tidy(sem.caus)$conf.high,
        
        sem.rcl.r.z = FisherZ(tidy(sem.rcl)$estimate),
        caus.rcl.r.z = FisherZ(tidy(caus.rcl)$estimate),
        sem.caus.r.z = FisherZ(tidy(sem.caus)$estimate))
      corr = rbind(corr,sub.corr)
      
      counter=counter+1
      print(counter)}
  }
}

sub.merge
```




